Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-21 checking-executions.tex
something

TODO REPHRASE THIS
While checking linearizability of a given execution is known to be NP-complete in general~\cite{journals/siamcomp/GibbonsK97} we show in this section that it is polynomial-time for data-differentiated executions of the priority queue. For pedagogical reasons, Section~\ref{ssec:seq_exec} introduces a polynomial-time recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into the language accepted by $LTS??$. The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. {\color {blue} Here we implicitly mix operation $o$ (when $\textit{call}_o(m,a)$, $\textit{ret}_o(m,b)$ in $e$) with $m(a,b)$.}

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
{\color{blue}When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure try to select a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value.} For instance, since $p_2$ is a maximal priority, it may select the operation $\textit{put}(d,p_2)$. {\color {blue}This operation is correct if $d$ is the last value with priority $p_2$,} and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
{\color {blue}Now, all the putted value that is not removed of priority $p_2$ do not exists, and the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations of priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. In this case, the value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, the only value of priority smaller than $p_2$ is $a$ and satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

@@

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. Thus, if $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set of values $D$ such that one of the following doesn't hold: $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$, {\color {blue}$\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ for all values that are unmatched and with maximal priority $x$, or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ for all maximal priority values $x$ where values of $x$ are all matched.} It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

@@

TODO UPDATE THE NOTATIONS IN THIS EXAMPLE

>>>>

%TODO REPHRASE THIS
%While checking linearizability of a given execution is known to be NP-complete in general~\cite{journals/siamcomp/GibbonsK97} we show in this section that it is polynomial-time for data-differentiated executions of the priority queue. 
We define a recursive procedure for checking linearizability of an execution w.r.t. $\seqPQ$.
To ease the exposition, Section~\ref{ssec:seq_exec} introduces a recursive procedure for checking whether a data-differentiated \emph{sequential} execution is admitted by the priority queue which is then extended to the concurrent case in Section~\ref{ssec:conc_exec}.

@@

Checking whether a data-differentiated sequential execution belongs to $\seqPQ$ could be implemented by checking membership into the language accepted by the LTS $PQ$. The recursive procedure $\textit{Check-PQ-Seq}$ outlined in Algorithm~\ref{alg:seq_check} is an alternative to this membership test. Roughly, it selects one or two operations in the input execution, checks whether their return values are correct by ignoring the order between the other operations other than how they are ordered w.r.t. the selected ones, and calls itself recursively on the execution without the selected operations. %{\color {blue} Here we implicitly mix operation $o$ (when $\textit{call}_o(m,a)$, $\textit{ret}_o(m,b)$ in $e$) with $m(a,b)$.}

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.} are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it may select the operation $\textit{put}(d,p_2)$. {This operation is correct if $d$ is the last value with priority $p_2$, and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
Since there is no other value of priority $p_2$ which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

@@

The following lemma states the correctness of $\textit{Check-PQ-Conc}$. Completeness follows easily from the properties of $\seqPQ$. Thus, if $\textit{Check-PQ-Conc}(e) = \textit{false}$, then there exists a set $D$ of values s.t. either  $\mathsf{EmptyRemove\text{-}Conc}(e \vert D)$ is false, or $\mathsf{UnmatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority that are not removed (and there exists at least one such value), or $\mathsf{MatchedMaxPriority\text{-}Conc}(e \vert D,x)$ is false for all the values $x$ of maximal priority (and these values are all removed in $e \vert D$). It can be easily seen that we get $e \vert D\not\sqsubseteq \seqPQ$ in all cases, which by the closure under projection of $\seqPQ$ implies, $e \not\sqsubseteq \seqPQ$ (since every linearization of $e$ includes as a subsequence a linearization of $e \vert D$).

@@

### 2017-04-21 co-regular-EPQ.tex
modify
something

The following result shows that deriving $\Gamma$-complete automata for each $\Gamma$ enables an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability. It is a direct consequence of the above definitions.

Section~\ref{ssec:aut} describes a $\mathsf{MatchedMaxPriority}$-complete automaton (TODO the other automata are discussed in Appendix~\ref{} or Icalp??) while Section~\ref{} discusses decidability results implied by the theorem above.

W.l.o.g., we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using the techniques introduced for regular FIFO queues in~\cite{DBLP:conf/icalp/BouajjaniEEH15} (see Appendix~\ref{} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

@@

@@

TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.

@@

{\color {blue}
Let $\mathcal{A}_{\textit{1-lar}}$ be the union of the four register automata mentioned above, then we prove that $\mathcal{A}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete. The detailed construction and proof can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.
}

\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete.
\end{restatable}

More details can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. %TODO MOVE THIS TO THE RELATED WORK

@@

The result of Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} still needs to be impove, since obtaining $y <_{\textit{pb}}^* x$ may requires arbitrary many intermediate values, and it is hard to store unbounded values by automata. Fortunately, we prove that the number of intermediate values is essentially bounded, as stated by the following lemma.

\begin{restatable}{lemma}{OBOrderHasBoundedLength}
\label{lemma:ob order has bounded length}
Given a data-differentiated execution $e$. Assume that $a <_{\textit{pb}} a_1 <_{\textit{pb}} \ldots <_{\textit{pb}} a_m <_{\textit{pb}} b$, then one of the following cases holds:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$ or $a <_{\textit{pb}}^C b$,

\item[-] $a <_{\textit{pb}}^A a_i <_{\textit{pb}}^B b$, or $a <_{\textit{pb}}^B a_i <_{\textit{pb}}^A b$, for some $i$.
%\item[-] $a <_{\textit{pb}}^B a_i <_{\textit{pb}}^A a_j <_{\textit{pb}}^B b$, for some $i$ and $j$,
\end{itemize}
\end{restatable}

By Lemma \ref{lemma:ob order has bounded length}, we can use register automata to detect $a <_{\textit{pb}}^* b$ by enumerating all possible enumerations of operations of $a$, $b$ and possibly $a_1$. Since some combinations of $<_{\textit{pb}}^*$ and gap-points are conflict, we finally reduce the number of potential enumerations of operations of $a$, $b$ and possibly $a_1$ into only five in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}.

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}, we construct a set $\textit{Auts}_{\textit{1-eq}}$ of register automata, and prove that $\mathsf{MatchedMaxPriority}^{=}$ is co-regular, as stated by the following lemma.

\begin{restatable}{lemma}{EPQOneEqualIsCoRegular}
\label{lemma:EPQ1Equal is co-regular}
$\mathsf{MatchedMaxPriority}^{=}$ is co-regular.
\end{restatable}

One such enumeration and its register automata is shown in \figurename~\ref{fig:an enumeration and its witness automaton}. $o$ is the rightmost gap-point of $b$, we rename the values that ``covers the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$'' (see Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}) into $d$, and rename the renaming values into $e$. In this figure, $c = \textit{call}(\textit{put},e,\top),\textit{ret}(\textit{put},e,\top)$, $\textit{call}(\textit{rm},e), \textit{ret}(\textit{rm},e),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$, $c_1 = c + \textit{call}(\textit{put},d,<r)$, $c_2 = c_1 + \textit{ret}(\textit{put},b,r)$, $c_3 = c_2 + \textit{ret}(\textit{rm},d)$, $c_4 = c + \textit{ret}(\textit{put},b,r) + \textit{ret}(\textit{rm},d)$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8 \textwidth]{figures/PIC-Enumeration-WitnessAutomata.pdf}
%\vspace{-10pt}
  \caption{An enumerations and its register automaton}
  \label{fig:an enumeration and its witness automaton}
\end{figure}

\subsection{Reduce into State Reachability}
\label{subsec:combine step-by-step linearizability and co-regular}

We also prove that $\mathsf{UnmatchedMaxPriority}$ are co-regular in Appendix \ref{subsec:appendix co-regular of EPQ2Lar}, Appendix \ref{subsec:appendix co-regular of EPQ2Equal}, and prove that $\mathsf{EmptyRemove}$ are co-regular in Appendix \ref{subsec:co-regular of EPQ3}. Then we can see that $\seqPQ$ is co-regular.

\begin{restatable}{lemma}{EPQIsCoRegular}
\label{lemma:EPQ is co-regular}
$\seqPQ$ is co-regular.
\end{restatable}

By Lemma \ref{lemma:EPQ as multi in MRpri for history} and Lemma \ref{lemma:EPQ is co-regular}, we can finally reduce the verification of linearizability w.r.t $\seqPQ$ into a reachability problem, as illustrated by the following theorem, where $\textit{Auts}_{\textit{EPQ}}$ is the set of witness automata of this section.

\begin{restatable}{theorem}{ReduceEPQIntoStateReachability}
\label{lemma:reduce EPQ into state reachability}
Given a data-independence implementation $\mathcal{I}$. $\mathcal{I} \sqsubseteq \seqPQ$, if and only if, $\mathcal{I} \cap \textit{Auts}_{\textit{EPQ}} = \emptyset$.
\end{restatable}

When the data domain and priority domain are both fixed to be finite set, the register automata can be simulated with finite automata, and by Lemma \ref{lemma:reduce EPQ into state reachability}, checking linearizable w.r.t $\seqPQ$ is decidable.

>>>>

Section~\ref{ssec:aut} describes a $\mathsf{MatchedMaxPriority}$-complete automaton, the other automata being defined in Appendix~\ref{} or Icalp??. Therefore, the following holds.

\begin{lemma}
There exists a $\Gamma$-complete automaton for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$.
\end{lemma}

When defining $\Gamma$-complete automata, we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using the techniques introduced for regular FIFO queues in~\cite{DBLP:conf/icalp/BouajjaniEEH15} (see Appendix~\ref{} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

@@

The following result shows that $\Gamma$-complete automata enable an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability. It is a direct consequence of the above definitions.

Section~\ref{subsec:combine step-by-step linearizability and co-regular} discusses decidability results implied by the theorem above.

%Section~\ref{ssec:aut} discusses

@@

%TODO WHY IT IS A SET OF AUTOMATA ? SAY WHAT ARE THE CASES THAT NEED TO BE COVERED. SAY WHAT IS THE CASE WE GIVE IN THE FIGURE. USE $\top$ AS MENTIONED AT THE BEGINNING OF THE SECTION INSTEAD OF 	$c$. USE BIG CAPS FOR SETS OF TRANSITION LABELS.
TODO COMPLETE THE FOLLOWING
Defining a $\mathsf{MatchedMaxPriority}^>$-complete automaton needs to consider ?? cases depending on the order between ??. For each case, we define an automaton recognizing the induced set of violations. For example, the automaton corresponding to the case ?? is pictured in in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. More details can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.

@@

%{\color {blue}
%Let $\mathcal{A}_{\textit{1-lar}}$ be the union of the four register automata mentioned above, then we prove that $\mathcal{A}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete. The detailed construction and proof can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}.
%}
%
%\begin{restatable}{lemma}{EPQOneLarisCoRegular}
%\label{lemma:EPQ1Lar is co-regular}
%$\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete.
%\end{restatable}

@@

To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$ using an automaton that tracks a bounded number of values, we show that the number of values needed to witness that $y <_{\textit{pb}}^* x$ for some $x$ and $y$ is bounded.
%
%While Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} doesn't imply a bound for 
%the number of values needs to witness that 
%
%The result of Lemma \ref{lemma:EPQ1Equal as pb order and gap-point} still needs to be impove, since obtaining $y <_{\textit{pb}}^* x$ may requires arbitrary many intermediate values, and it is hard to store unbounded values by automata. Fortunately, we prove that the number of intermediate values is essentially bounded, as stated by the following lemma.

\begin{restatable}{lemma}{OBOrderHasBoundedLength}
\label{lemma:ob order has bounded length}
Let $e$ be a data-differentiated execution such that $a <_{\textit{pb}} a_1 <_{\textit{pb}} \ldots <_{\textit{pb}} a_m <_{\textit{pb}} b$ holds for some set of values $a$, $a_1$,$\ldots$,$a_m$, $b$. Then, one of the following holds:
\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$, or $a <_{\textit{pb}}^C b$,

\item[-] $a <_{\textit{pb}}^A a_i <_{\textit{pb}}^B b$ or $a <_{\textit{pb}}^B a_i <_{\textit{pb}}^A b$, for some $i$.
%\item[-] $a <_{\textit{pb}}^B a_i <_{\textit{pb}}^A a_j <_{\textit{pb}}^B b$, for some $i$ and $j$,
\end{itemize}
\end{restatable}

To characterize violations to linearizability w.r.t. $\mathsf{MatchedMaxPriority}^{=}$, one has to consider all the possible orders between call/return actions of the operations on values $a$, $b$, and $a_i$ in Lemma \ref{lemma:ob order has bounded length}, and the right-most gap point of $b$. Excluding the inconsistent cases, we are left with 5 possible orders that are listed in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}. For each case, we define an automaton recognizing the induced set of violations. For instance, one such case and the corresponding register automaton is shown in \figurename~\ref{fig:an enumeration and its witness automaton} ($o$ denotes the rightmost gap-point of $b$). In this case, the conditions in Lemma~\ref{lemma:EPQ1Equal as pb order and gap-point} are equivalent to the fact that intuitively, the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$ is covered by lower priority values (and thus, there is no gap-point of $b$ which occurs after $\textit{call}(\textit{rm},a)$). Using again the data-independence property, these lower priority values can all be renamed to a fixed value $d$, and the other values to a fixed value $e$.

%We denote by $o$ the rightmost gap-point of $b$, we rename the values that ``cover the time interval from $\textit{call}(\textit{rm},a)$ to $\textit{ret}(\textit{rm},b)$'' (see Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}) into $d$, and rename the renaming values into $e$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8 \textwidth]{figures/PIC-Enumeration-WitnessAutomata.pdf}
%\vspace{-10pt}
  \caption{A case in deriving a $\mathsf{MatchedMaxPriority}^{=}$-complete automaton. We use the following notations: $c = \textit{call}(\textit{put},e,\top),\textit{ret}(\textit{put},e,\top)$, $\textit{call}(\textit{rm},e), \textit{ret}(\textit{rm},e),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$, $c_1 = c + \textit{call}(\textit{put},d,<r)$, $c_2 = c_1 + \textit{ret}(\textit{put},b,r)$, $c_3 = c_2 + \textit{ret}(\textit{rm},d)$, $c_4 = c + \textit{ret}(\textit{put},b,r) + \textit{ret}(\textit{rm},d)$.}
  \label{fig:an enumeration and its witness automaton}
\end{figure}

%we can use register automata to detect $a <_{\textit{pb}}^* b$ by enumerating all possible enumerations of operations of $a$, $b$ and possibly $a_1$. Since some combinations of $<_{\textit{pb}}^*$ and gap-points are conflict, we finally reduce the number of potential enumerations of operations of $a$, $b$ and possibly $a_1$ into only five in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}.

%In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}, we construct a set $\textit{Auts}_{\textit{1-eq}}$ of register automata, and prove that $\mathsf{MatchedMaxPriority}^{=}$ is co-regular, as stated by the following lemma.
%
%\begin{restatable}{lemma}{EPQOneEqualIsCoRegular}
%\label{lemma:EPQ1Equal is co-regular}
%$\mathsf{MatchedMaxPriority}^{=}$ is co-regular.
%\end{restatable}

\subsection{Decidability Result}
\label{subsec:combine step-by-step linearizability and co-regular}

We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method manipulates a finite number of local variables which store Boolean values, or data values from $\mathbb{D}$. Methods communicate through a finite number of shared variables that also store Boolean values, or data values from $\mathbb{D}$. Data values may be assigned, but never used in program predicates (e.g., in the conditions of if-then-else and while statements) so as to ensure data independence. This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. Since the $\Gamma$-complete automata $A(\Gamma)$ we define use a fixed set $D=\{a,b,d,e,\top\}$ of values, we have that $\mathcal{I}\cap A(\Gamma)\neq\emptyset$ for some $\Gamma$ iff $\mathcal{I}_D\cap A(\Gamma)\neq\emptyset$ where $\mathcal{I}_D$ is the subset of $\mathcal{I}$ that uses only values in $D$. The set of executions $\mathcal{I}_D$ can be represented by a Petri Net since bounding the set of values allows us to represent each thread with a finite-state machine. For a fixed set of priorities $P$, the register automata $A(\Gamma)$ can be transformed to finite-state automata since the number of possible valuations of the registers is bounded.

\begin{theorem}
Verifying whether an implementation in $\mathcal{C}$ is linearizable w.r.t. $\seqPQ$ is PSPACE-complete for a fixed number of threads, and EXPSPACE-complete otherwise.
\end{theorem}

%We also prove that $\mathsf{UnmatchedMaxPriority}$ are co-regular in Appendix \ref{subsec:appendix co-regular of EPQ2Lar}, Appendix \ref{subsec:appendix co-regular of EPQ2Equal}, and prove that $\mathsf{EmptyRemove}$ are co-regular in Appendix \ref{subsec:co-regular of EPQ3}. Then we can see that $\seqPQ$ is co-regular.
%
%\begin{restatable}{lemma}{EPQIsCoRegular}
%\label{lemma:EPQ is co-regular}
%$\seqPQ$ is co-regular.
%\end{restatable}
%
%By Lemma \ref{lemma:EPQ as multi in MRpri for history} and Lemma \ref{lemma:EPQ is co-regular}, we can finally reduce the verification of linearizability w.r.t $\seqPQ$ into a reachability problem, as illustrated by the following theorem, where $\textit{Auts}_{\textit{EPQ}}$ is the set of witness automata of this section.
%
%\begin{restatable}{theorem}{ReduceEPQIntoStateReachability}
%\label{lemma:reduce EPQ into state reachability}
%Given a data-independence implementation $\mathcal{I}$. $\mathcal{I} \sqsubseteq \seqPQ$, if and only if, $\mathcal{I} \cap \textit{Auts}_{\textit{EPQ}} = \emptyset$.
%\end{restatable}
%
%When the data domain and priority domain are both fixed to be finite set, the register automata can be simulated with finite automata, and by Lemma \ref{lemma:reduce EPQ into state reachability}, checking linearizable w.r.t $\seqPQ$ is decidable.

### 2017-04-21 introduction.tex
something

Next, we show that checking whether an execution violates this ``local'' property for a value $x$ can be done using a class of register automata~\cite{DBLP:journals/tcs/KaminskiF94,DBLP:conf/icalp/Cerans94,DBLP:conf/stacs/SegoufinT11} (transition systems where the states consist of a fixed set of registers that can receive values and be compared). Actually, only two registers are needed: one register $r_1$ for storing a priority guessed at the initial state, and one register $r_2$ for reading priorities as they occur in the execution and comparing them with the one stored in $r_1$. We show that registers storing values added to or removed from the priority queue are not needed, since any data-independent implementation admits a violation to linearizability whenever it admits a violation where the number of values is constant, and at most 4 (the number of priorities can still be unbounded).

>>>>

Next, we show that checking whether an execution violates this ``local'' property for a value $x$ can be done using a class of register automata~\cite{DBLP:journals/tcs/KaminskiF94,DBLP:conf/icalp/Cerans94,DBLP:conf/stacs/SegoufinT11} (transition systems where the states consist of a fixed set of registers that can receive values and be compared). Actually, only two registers are needed: one register $r_1$ for storing a priority guessed at the initial state, and one register $r_2$ for reading priorities as they occur in the execution and comparing them with the one stored in $r_1$. We show that registers storing values added to or removed from the priority queue are not needed, since any data-independent implementation admits a violation to linearizability whenever it admits a violation where the number of values is constant, and at most 5 (the number of priorities can still be unbounded).

### 2017-04-21 priority-queues.tex
something

Besides data independence, the sequential behaviors of the priority queue satisfy the following closure property: a behavior remains valid when removing all the operations with an argument in some set of values $D \subseteq \mathbb{D}$ and any $\textit{rm}(\textit{empty})$ operation (since they are read-only and they don't affect the queue's state).
In order to distinguish between different $\textit{rm}(\textit{empty})$ operations while simplifying the technical exposition, we assume that they receive as argument a value, i.e., call actions are of the form $\textit{call}_o(\textit{rm},\textit{empty},a)$ for some $a\in \mathbb{D}$. We will make explicit this argument only when needed in our technical development. The projection $e \vert D$ of an execution $e$ to a set of values $D \subseteq \mathbb{D}$ is obtained from $e$ by erasing all call/return actions with an argument not in $D$. We write $e \setminus x$ for the projection $e \vert_{ \mathbb{D} \setminus \{ x \} }$. Let $\textit{proj}(e)$ be the set of all projections of $e$ to a set of values. {\color {blue} The proof of the following lemma can be found in Appendix \ref{sec:appendix proofs in section priority queue and data-independence}.}

@@

The following lemma states that by data-independence, it is enough to consider only data-differentiated executions when checking linearizability. {\color {blue}It is proved in Appendix \ref{sec:appendix proofs in section priority queue and data-independence}.} This is similar to that in \cite{conf/tacas/AbdullaHHJR13,DBLP:conf/icalp/BouajjaniEEH15}, where they use the notion of data-independence in \cite{conf/popl/Wolper86}.
Section~\ref{sec:checking inclusion by recursive procedure} will focus on characterizing linearizability for data-differentiated executions. %The proof of this lemma can be found in Appendix \ref{sec:appendix in section data-independence of EPQ}.

>>>>

Besides data independence, the sequential behaviors of the priority queue satisfy the following closure property: a behavior remains valid when removing all the operations with an argument in some set of values $D \subseteq \mathbb{D}$ and any $\textit{rm}(\textit{empty})$ operation (since they are read-only and they don't affect the queue's state).
In order to distinguish between different $\textit{rm}(\textit{empty})$ operations while simplifying the technical exposition, we assume that they receive as argument a value, i.e., call actions are of the form $\textit{call}_o(\textit{rm},\textit{empty},a)$ for some $a\in \mathbb{D}$. We will make explicit this argument only when needed in our technical development. The projection $e \vert D$ of an execution $e$ to a set of values $D \subseteq \mathbb{D}$ is obtained from $e$ by erasing all call/return actions with an argument not in $D$. We write $e \setminus x$ for the projection $e \vert_{ \mathbb{D} \setminus \{ x \} }$. Let $\textit{proj}(e)$ be the set of all projections of $e$ to a set of values. The proof of the following lemma can be found in Appendix \ref{sec:appendix proofs in section priority queue and data-independence}.

@@

The following lemma states that by data-independence, it is enough to consider only data-differentiated executions when checking linearizability (see Appendix \ref{sec:appendix proofs in section priority queue and data-independence}). This is similar to that in \cite{conf/tacas/AbdullaHHJR13,DBLP:conf/icalp/BouajjaniEEH15}, where they use the notion of data-independence in \cite{conf/popl/Wolper86}.
Section~\ref{sec:checking inclusion by recursive procedure} will focus on characterizing linearizability for data-differentiated executions. %The proof of this lemma can be found in Appendix \ref{sec:appendix in section data-independence of EPQ}.
